Paper 1: Triangulating trends in inequality around the world from secondary databases

### 2023-04-05 PhD_pages/Paper_1/Presentation_INET_16_feb_inequality_around_the_world.qmd
Paper 1: drafting and testing out workflow for tables

---
title: "Triangulating a global view on inequality"
author: "<br> <br> Joe Hasell – DPhil candidate at DSPI University of Oxford; Our World in Data<br> <br> Pablo Arriagada – Our World in Data"
format: 
    revealjs:
        theme: [default, revealjs_joe_theme.scss] 
bibliography: references.bib
---

>>>>

---
title: "Triangulating a global view on inequality"
author: "<br> <br> Joe Hasell – DPhil candidate at DSPI University of Oxford; Our World in Data<br> <br> Pablo Arriagada – Our World in Data"
format: 
    revealjs:
        theme: [default, revealjs_joe_theme.scss] 
bibliography: references_old.bib
execute:
  eval: false
---

### 2023-04-05 PhD_pages/Paper_1/paper1.qmd
Paper 1: drafting and testing out workflow for tables

---
title: "Paper 1: Triangulating trends in inequality around the world from secondary databases"
format: html
bibliography: references_07_03_2023.bib
execute:
  echo: false
  warning: false
---

@@

In part at least, this increased concern reflects an increased availability of distributional data. New and expanded sources of data have shed new light on extent of inequality around the world and how it is changing. This has enabled concerns about high or rising inequality to be expressed in new ways, formulated more precisely and subsequently monitored. The supply and demand for global data on inequality have been mutually reinforcing.

Two sources of data have been particularly impactful.

One is the collection of national household survey data brought together by the World Bank, published online through its *Poverty and Inequality Platform*. This is produced as part of the World Bank's remit to monitor global poverty and shared prosperity. Most extensive harmonized . Its impact lies in its extensive use also by the UN in setting and monitoring many of the Sustainable Development Goals.

Another especially important source is the World Inequality Database.

 Capital in the 21st C -> WID. "The 1 %".

Both datasets have continued to improve adn expand. In the case of PIP, including high income countries, following recommendations of the Atkinson commission. In the case of WID – transition from earlier, using with more minimal requirements in terms of source data.

In terms of data availability we are thus in a better position than ever to take a global viw on inequlity. Yet amidst apparent wealth of data, limited understanding of very basics.

Contradictory messages emerge concerning the most basic descriptions from the two databases.

The view from *World Inequality Database* and its associated global report is that inequality is rising in 'nearly everywhere' [@alvaredo_world_2018; @chancel_world_2021].[^WID_quotes] The World Bank data and its biannual report offers a much more mixed picture – in which inequality is rising and falling in roughly equal numbers of countries.

@@

This stark difference is reflected in decompositions of global inequality into between- and within-country components. (Explain)

Thus despite the great data, there remains fundamental ambiguity surrounding which of two possible worlds we live in. In one world, rising inequality within countries has been an almost universal phenomenon, and has now come to dominate income differences between countries. In another, within-country inequality trends in recent decades have been mixed, and the differences in average incomes between countries remain the lion's share of the globally.

Which of these worlds we live in matters hugely – both in our conception, and of where to place efforts to combat inequality in the future.

The first world suggests the winners from globalization and economic growth overwhelmingly and almost universally, have been a global elite. It situates inquality within countries, but emphasises global factors – including fundamental laws of capitalism with regard to capital accumulation (Piketty) – and the need for international coordination over and above national policy.

The second world and the importance of access to both produced and human capital , national policy (Ravallion, 2018; Corlett, 2016).

### Unpicking the two worlds

Many methodolodical choices concerning underlying data sources, the constraints on temporal coverage this gives rise to,  adjust, conception of inequality. These to some extent come bundled together when making the choice between PIP and WID. That is to say, the empirical question of 'world' of inequality we live in depends on which 'world' of inequality *measurement* we adopt.

Whilst recommending particular aspects, for particular countries, the choice put before us requires wholesale adoption of a bundle whose relative strengths and weaknesses when applied globally is not well understood.

@@

In this paper, I focus on trends in within-country inequality in recent decades.(In a companion paper, I look at the global distribution.) I line up what I can. I triangulate what we know – . And Identify what we don't know – the remaining differences – and try to pinpoint where the differences lie.

@@

@@

## Data

### PIP

### WID

### LIS

## Methods

@@

>>>>

---
title: "Paper 1: Triangulating trends in inequality around the world from secondary databases"
format: html
bibliography: references_export_2023-04-04.bib
execute:
  echo: false
  warning: false
---

@@

In part at least, this increased concern reflects an increase in the *availability* of distributional data. New and expanded sources of data have shed new light on extent of inequality around the world and how it is changing. This has enabled concerns about high or rising inequality to be expressed in new ways, formulated more precisely and subsequently monitored. The supply and demand for global data on inequality have been mutually reinforcing.

Two sources of data have been particularly impactful in this regard: the World Bank's harmonized collection of household survey data, published online through its [*Poverty and Inequality Platform (PIP)*](https://pip.worldbank.org/); and the collection of inequality estimates – most notably top income shares – brought together at the [*World Inequality Database (WID)*](https://wid.world/).^[EDIT: PIP used to be Povcal. It draws upon the work of various regional teams. WID used to be WTID.]

These two sources are the only regularly-updated databases providing global coverage on a relatively consistent basis.^[EDIT: Both datasets suffer important limitations with respect to comparability acrosss time and location. These will be discussed in the data section below. Both datasets, however, take significant steps to harmonize, *ex-post*, the available underlying data. In this respect they can be distinguished from databases such as WIID and SWIID [@atkinson_promise_2001]] Both have played substantial agenda-setting roles in academic research, policy and public debate.

Given its extensive coverage of low- and middle-income countries, the World Bank data occupies a central place in international development policy and research. Most notably, the data is relied on by the UN to set and monitor progress towards many of the targets included within Sustainable Development Goals. It has also to date underpinned the majority of academic studies on inequality that take a global view.^[[Including @Kanbur2022-lk; @lakner_global_2016; @anand_global_2015; @Nino-Zarazua2017-tv; @Ravallion2014-mk] TODO: Check and more refs].

The *WID* database emerged as institutional focal-point to host and coordinate inequality estimates produced as part of the burgeoning literature on top incomes [@atkinson_top_2014; @piketty_capital_2014; @alvaredo_top_2013]. This development of this data has opened new avenues for the study of inequality right across the social sciences (TODO: example refs), underpinning a significant re-evaluation of its causes and consequences [@alvaredo_top_2013]. It has also surely had a substantial impact on the public's conception of inequality, contributing to the momentum generated by the Occupy movement and its framing of inequality in terms of 'the 1%' vs the '99%''.

#### Two worlds of inequality measurement

The two . They in some sense represent *two different worlds* of inequality measurement.

In terms of methods, source, conception of inequality and research questions. "The 1 %" – pretax, with many assumptions needed to get to national income. These come as a bundle. Although both institutions continue to take many steps to provide transparency and reproducibility to their estimates, they remain a black box.

Also in the research questions the datasets aim at. Capital vs labour share. .– including fundamental laws of capitalism with regard to capital accumulation (Piketty) –

Most crucially, the worlds also diverge in terms of headline messages as to inequality around the world. Contradictory messages emerge concerning the most basic descriptions from the two databases.

They differ as to how ubiquitous a phenomenon rising inequality is around the world. The view from *World Inequality Database* and its associated global report is that inequality is rising in 'nearly everywhere' [@alvaredo_world_2018; @chancel_world_2021].[^WID_quotes] In contrast, the World Bank data and its biannual report offers a much more mixed picture – in which inequality is rising and falling in roughly equal numbers of countries.

@@

Secondly, the two worlds differ in their conception of global inequality....  decompositions of global inequality into between- and within-country components. (Explain

Incredible efforts, involving the work of hundreds of researchers. Both datasets have continued to improve adn expand. In terms of data availability we are thus in a better position than ever to take a global viw on inequlity. And yet, amidst this wealth of data, limited understanding of very basics concerning inequality around the world.

There remains fundamental ambiguity surrounding which of two possible worlds we live in.

The '*WID* world' rising inequality within countries has been an almost universal phenomenon. suggests the winners from globalization and economic growth overwhelmingly have been a global elite. It sees the bulk of global inequalities as lying within nations, but emphasises global factors in terms of causes and hence the need for international coordination to combat inequality.

The '*PIP* world' – within-country inequality trends in recent decades have been mixed, and the differences in average incomes between countries remain the lion's share of the globally.As such the importance of national policy (Ravallion, 2018; Corlett, 2016). access to both produced and human capital.

Knowing which of these worlds we live in matters a great deal – for our understanding of inequality today, how it has changed, and also where efforts to combat inequality in the future are likely to be most successful.

Many methodolodical choices concerning underlying data sources, the constraints on temporal coverage this gives rise to,  adjust, conception of inequality. These to some extent come bundled together when making the choice between PIP and WID.

In many particular contexts There are known strengths and weaknesses. However in using the datasets a global view on inequality, requires choosing between methodological 'bundles' whose relative strengths and weaknesses, applied globally, are not well understood.

### Connecting the two worlds
*(In this section I will explain the plan for the paper)*

@@

In this paper, I focus on trends in within-country inequality in recent decades.(In a companion paper, I look at the global distribution.) I line up what I can. I triangulate what we know – What is true irrespective of which 'world' we consider. And Identify what we don't know – the remaining differences – and try to pinpoint where the differences lie.

@@

Explain approach. (Including LIS – which is close to PIP world (indeed PIP takes data from LIS for a number of high income countries.))

@@

## Methods and data

@@

### Data

#### PIP

#### WID

#### LIS

#### Coverage and lining-up observations

## Results

### Comparison of typical indicators

- Relative change WID 1% and PIP Gini

### Aligning indices

- Show that measure doesn't matter for WID until top 1 (check Palma ratio)

- Check out how much different metrics matter for PIP.

-----
